VariantSpark: Population Scale Clustering of Genotype Information

### 2015-11-25 doc/publications/journal2014/paper.tex
updates for reviewers

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for the complete genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

@@

@@

@@

\label{scalingcomparison}
\begin{table}[h!]
\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
      \begin{tabular}{|c|ccc|ccc|c|c|}
        \hline
           Portion &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering}  \\
& executors & memory & time  & executors & memory & time \\
  \hline
20\%		& 64	& 2	& 11min 53sec	& 64	& 6	& 1hr 10min	\\
40\%		& 64	& 2	& 19min 9sec	& 64	& 12	& 2hr 19min	\\
60\%		& 64	& 2	& 26min 34sec	& 64	& 17	& 3hr 33min	\\
100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
  \hline
      \end{tabular}
\end{table}

>>>>

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12, 19, 27 and 40 minutes for 20, 40, 60 and 100 percent of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for the complete genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

@@

To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants. For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.

@@

%Compare with our old implementation still? Or just have the VariantAnnotation one.
%Do we acknowledge that this is a reviewer suggestion in any way?
We also compare this approach to one using the `VariantAnnotation' package. Reading the VCF file to a matrix with `readGT' is faster than reading a VCF file to a table using R's `read.table', at 116 seconds compared to 207 seconds,
however, because `readGT' only includes genotype information in the matrix, we can simplify our function that converts this information into numeric values. This speeds up processing time greatly. More implementation details are available
in the supplementary material.

@@

\label{scalingcomparison}
\begin{table}[h!]
\caption{The resources consumption on different subsets of the entire autosome (chromosomes 1-22). Memory specified is the memory allocated to each executor.}
      \begin{tabular}{|c|ccc|ccc|c|c|}
        \hline
           Portion &  \multicolumn{3}{ c |}{Pre-processing} & \multicolumn{3}{ c |}{Clustering}  \\
& executors & memory & time  & executors & memory & time \\
  \hline
20\%		& 64	& 2	& 11min 53sec	& 64	& 6	& 1hr 10min	\\
40\%		& 64	& 2	& 19min 9sec	& 64	& 12	& 2hr 19min	\\
60\%		& 64	& 2	& 26min 34sec	& 64	& 17	& 3hr 33min	\\
100\%	& 64	& 2	& 40min 48sec	& 40	& 24	& 14hr 44min	\\
???\%	& 64 & 2	& ??min ??sec	& 40 & 24	& 27hr 46min	\\
  \hline
      \end{tabular}
\end{table}
